Burning sage: Reversing the curse of dimensionality in the visualization of high-dimensional data

### 2021-07-26 paper/paper.Rmd
Finishing second revision.

This affects many aspects of data analysis: minimizing the error during model fitting relies on effective optimization techniques, non-parametric modeling requires finding nearest neighbors which may be far away and sampling from high-dimensional distributions is likely to have points far from the population mean. @Donoho00 considers the curse of dimensionality as a blessing, because the sparsity can be leveraged for computational efficiency. This is used in regularization methods, like lasso, to penalize model complexity. The penalty term results in shrinking (some of) the parameter estimates towards zero.

@@

Common solutions to overplotting include transparency and binned or density displays [@chang]. These are designed to tackle large sample size, rather than dimension reduction problems. A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection. These approaches are still useful for large sample sizes, and could be applied after the transformation.

@@

Figure \ref{fig:sketch} illustrates the comparison to be made, using something we can easily picture, a 3D sphere. Projecting the data from within a 3D sphere to 2D (grey disk) will result in mass being condensed into the disk. Imagine comparing the volume of a cylinder at different locations in the disk. A centered cylinder has more volume. This is exaggerated as $p$ increases: the centered cylinder has much more volume than any other cylinder.

@@

Before applying the transformation, it can be important to pre-process the data, as is generally required with methods for the visualisation of high-dimensional data. In particular, when working with projections that combine multiple variables it is important to ensure a common scale across the dimensions. This is especially important with the sage display, where the overall scale (set by the maximum radius $R$) determines the effect of the transformation. Scaling is commonly also combined with centering as a pre-processing step. This is less important in our implementation because we always center the projected data before applying our transformation (i.e. the projected mean defines the origin of our polar coordinate system in the plane). Centering and scaling together would be achieved by standardizing each variable in the data to be centered at 0 and have a variance equal to 1.

We may also wish to apply non-linear transformations before visualising the data or combining variables in a projection. One example would be to use a log transformation in the case of a variable that has a skewed distribution. These types of transformations will generally change the distribution across the high-dimensional volume (and in particular also the position of the mean), and can thus make the sage display more useful for a wide range of observed data distributions.

@@

We next use the new display to look at different distributions of images from the Google QuickDraw collection [@quickdraw-paper]. These are $28\times28=784$ pixel grey scale data that are available publicly. In this example, we sample 1000 images from three types of sketches (banana, cactus, crab) and see if we can separate the classes in the high-dimensional parameter space.

@@

An alternative is the slice tour [@sliceTour] which  allows distributions
of points around the center of the data to be explored using sections instead of projections.
Instead of adjusting for effects from projecting a high-dimensional volume, the solution with the slice tour is to only highlight projected points if they are within a small volume around the current projection plane, thus showing a local view of the distribution.
The slice tour is useful when there are large numbers of observations or if there is concave structure in the data.
Note that only few methods are currently available for looking at sections of the data in a systematic fashion, and further work is needed to fully explore the potential of this type of visualisation. In smaller datasets we may also use interactive approaches such as linked brushing to explore for local structures or conditional features.

@@

The authors gratefully acknowledge the support of the Australian Research Council. The paper was written in `rmarkdown` [@rmarkdown] using `knitr` [@knitr].

>>>>

This affects many aspects of data analysis: minimizing the error during model fitting relies on effective optimization techniques, non-parametric modeling requires finding nearest neighbors which may be far away and sampling from high-dimensional distributions is likely to have points far from the population mean. Although not directly relevant for the topic of this paper, it is interesting to note that @Donoho00 considers the curse of dimensionality as a blessing, because the sparsity can be leveraged for computational efficiency. This is used in regularization methods, like lasso, to penalize model complexity. The penalty term results in shrinking (some of) the parameter estimates towards zero.

@@

Common solutions to overplotting include transparency and binned or density displays [@carr87], for a practical and accessible discussion see Section 5.5 of @chang. These are designed to tackle large sample size, rather than dimension reduction problems. A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection. These approaches are still useful for large sample sizes, and could be applied after the transformation.

@@

Figure \ref{fig:sketch} illustrates the comparison to be made, using something we can easily picture, a 3D sphere. Projecting the data from within a 3D sphere to 2D (gray disk) will result in mass being condensed into the disk. Imagine comparing the volume of a cylinder at different locations in the disk. A centered cylinder has more volume. This is exaggerated as $p$ increases: the centered cylinder has much more volume than any other cylinder.

@@

Before applying the transformation, it can be important to pre-process the data, as is generally required with methods for the visualization of high-dimensional data. In particular, when working with projections that combine multiple variables it is important to ensure a common scale across the dimensions. This is especially important with the sage display, where the overall scale (set by the maximum radius $R$) determines the effect of the transformation. Scaling is commonly also combined with centering as a pre-processing step. This is less important in our implementation because we always center the projected data before applying our transformation (i.e. the projected mean defines the origin of our polar coordinate system in the plane). Centering and scaling together would be achieved by standardizing each variable in the data to be centered at 0 and have a variance equal to 1.

We may also wish to apply non-linear transformations before visualizing the data or combining variables in a projection. One example would be to use a log transformation in the case of a variable that has a skewed distribution. These types of transformations will generally change the distribution across the high-dimensional volume (and in particular also the position of the mean), and can thus make the sage display more useful for a wide range of observed data distributions.

@@

We next use the new display to look at different distributions of images from the Google QuickDraw collection [@quickdraw-paper]. These are $28\times28=784$ pixel gray scale data that are available publicly. In this example, we sample 1000 images from three types of sketches (banana, cactus, crab) and see if we can separate the classes in the high-dimensional parameter space.

@@

\textcolor{red}{We have motivated the transformation from geometric arguments: equal volume of the high-dimensional hypersphere should be mapped onto equal area in the two-dimensional projection plane. We could also think about this is in terms of the distribution function: a uniform distribution in high-dimensions should still look uniform in the projected view. Our approach is defining the transformation assuming that points are uniformly distributed in the hypersphere. This could be generalized to work with the empirical radial cumulative distribution function instead.}

An alternative is the slice tour [@sliceTour] which  allows distributions
of points around the center of the data to be explored using sections instead of projections.
Instead of adjusting for effects from projecting a high-dimensional volume, the solution with the slice tour is to only highlight projected points if they are within a small volume around the current projection plane, thus showing a local view of the distribution.
The slice tour is useful when there are large numbers of observations or if there is concave structure in the data.
Note that only few methods are currently available for looking at sections of the data in a systematic fashion, and further work is needed to fully explore the potential of this type of visualization. In smaller datasets we may also use interactive approaches such as linked brushing to explore for local structures or conditional features.

@@

The authors gratefully acknowledge the support of the Australian Research Council. The paper was written in `rmarkdown` [@rmarkdown] using `knitr` [@knitr]. While the paper title was originally derived from a pun, we thank the reviewer who suggested that it could also be considered an acronym: BuRniNG SaGE: Baseline Radial Non-Linear Geometric Scaling for Graphical Exploration.

### 2021-07-26 paper/paper.tex
Finishing second revision.

This affects many aspects of data analysis: minimizing the error during
model fitting relies on effective optimization techniques,
non-parametric modeling requires finding nearest neighbors which may be
far away and sampling from high-dimensional distributions is likely to
have points far from the population mean. \citet{Donoho00} considers the
curse of dimensionality as a blessing, because the sparsity can be
leveraged for computational efficiency. This is used in regularization
methods, like lasso, to penalize model complexity. The penalty term
results in shrinking (some of) the parameter estimates towards zero.

@@

Common solutions to overplotting include transparency and binned or
density displays \citep{chang}. These are designed to tackle large
sample size, rather than dimension reduction problems. A low-dimensional
projection represented as a scatterplot, binned plot, or contour plot,
will still suffer from data piling and reduced resolution near the
center of the projection. These approaches are still useful for large
sample sizes, and could be applied after the transformation.

@@

Figure \ref{fig:sketch} illustrates the comparison to be made, using
something we can easily picture, a 3D sphere. Projecting the data from
within a 3D sphere to 2D (grey disk) will result in mass being condensed
into the disk. Imagine comparing the volume of a cylinder at different
locations in the disk. A centered cylinder has more volume. This is
exaggerated as \(p\) increases: the centered cylinder has much more
volume than any other cylinder.

@@

Before applying the transformation, it can be important to pre-process
the data, as is generally required with methods for the visualisation of
high-dimensional data. In particular, when working with projections that
combine multiple variables it is important to ensure a common scale
across the dimensions. This is especially important with the sage
display, where the overall scale (set by the maximum radius \(R\))
determines the effect of the transformation. Scaling is commonly also
combined with centering as a pre-processing step. This is less important
in our implementation because we always center the projected data before
applying our transformation (i.e.~the projected mean defines the origin
of our polar coordinate system in the plane). Centering and scaling
together would be achieved by standardizing each variable in the data to
be centered at 0 and have a variance equal to 1.

We may also wish to apply non-linear transformations before visualising
the data or combining variables in a projection. One example would be to
use a log transformation in the case of a variable that has a skewed
distribution. These types of transformations will generally change the
distribution across the high-dimensional volume (and in particular also
the position of the mean), and can thus make the sage display more
useful for a wide range of observed data distributions.

@@

We next use the new display to look at different distributions of images
from the Google QuickDraw collection \citep{quickdraw-paper}. These are
\(28\times28=784\) pixel grey scale data that are available publicly. In
this example, we sample 1000 images from three types of sketches
(banana, cactus, crab) and see if we can separate the classes in the
high-dimensional parameter space.

@@

An alternative is the slice tour \citep{sliceTour} which allows
distributions of points around the center of the data to be explored
using sections instead of projections. Instead of adjusting for effects
from projecting a high-dimensional volume, the solution with the slice
tour is to only highlight projected points if they are within a small
volume around the current projection plane, thus showing a local view of
the distribution. The slice tour is useful when there are large numbers
of observations or if there is concave structure in the data. Note that
only few methods are currently available for looking at sections of the
data in a systematic fashion, and further work is needed to fully
explore the potential of this type of visualisation. In smaller datasets
we may also use interactive approaches such as linked brushing to
explore for local structures or conditional features.

@@

The authors gratefully acknowledge the support of the Australian
Research Council. The paper was written in \texttt{rmarkdown}
\citep{rmarkdown} using \texttt{knitr} \citep{knitr}.

>>>>

This affects many aspects of data analysis: minimizing the error during
model fitting relies on effective optimization techniques,
non-parametric modeling requires finding nearest neighbors which may be
far away and sampling from high-dimensional distributions is likely to
have points far from the population mean. Although not directly relevant
for the topic of this paper, it is interesting to note that
\citet{Donoho00} considers the curse of dimensionality as a blessing,
because the sparsity can be leveraged for computational efficiency. This
is used in regularization methods, like lasso, to penalize model
complexity. The penalty term results in shrinking (some of) the
parameter estimates towards zero.

@@

Common solutions to overplotting include transparency and binned or
density displays \citep{carr87}, for a practical and accessible
discussion see Section 5.5 of \citet{chang}. These are designed to
tackle large sample size, rather than dimension reduction problems. A
low-dimensional projection represented as a scatterplot, binned plot, or
contour plot, will still suffer from data piling and reduced resolution
near the center of the projection. These approaches are still useful for
large sample sizes, and could be applied after the transformation.

@@

Figure \ref{fig:sketch} illustrates the comparison to be made, using
something we can easily picture, a 3D sphere. Projecting the data from
within a 3D sphere to 2D (gray disk) will result in mass being condensed
into the disk. Imagine comparing the volume of a cylinder at different
locations in the disk. A centered cylinder has more volume. This is
exaggerated as \(p\) increases: the centered cylinder has much more
volume than any other cylinder.

@@

Before applying the transformation, it can be important to pre-process
the data, as is generally required with methods for the visualization of
high-dimensional data. In particular, when working with projections that
combine multiple variables it is important to ensure a common scale
across the dimensions. This is especially important with the sage
display, where the overall scale (set by the maximum radius \(R\))
determines the effect of the transformation. Scaling is commonly also
combined with centering as a pre-processing step. This is less important
in our implementation because we always center the projected data before
applying our transformation (i.e.~the projected mean defines the origin
of our polar coordinate system in the plane). Centering and scaling
together would be achieved by standardizing each variable in the data to
be centered at 0 and have a variance equal to 1.

We may also wish to apply non-linear transformations before visualizing
the data or combining variables in a projection. One example would be to
use a log transformation in the case of a variable that has a skewed
distribution. These types of transformations will generally change the
distribution across the high-dimensional volume (and in particular also
the position of the mean), and can thus make the sage display more
useful for a wide range of observed data distributions.

@@

We next use the new display to look at different distributions of images
from the Google QuickDraw collection \citep{quickdraw-paper}. These are
\(28\times28=784\) pixel gray scale data that are available publicly. In
this example, we sample 1000 images from three types of sketches
(banana, cactus, crab) and see if we can separate the classes in the
high-dimensional parameter space.

@@

\textcolor{red}{We have motivated the transformation from geometric arguments: equal volume of the high-dimensional hypersphere should be mapped onto equal area in the two-dimensional projection plane. We could also think about this is in terms of the distribution function: a uniform distribution in high-dimensions should still look uniform in the projected view. Our approach is defining the transformation assuming that points are uniformly distributed in the hypersphere. This could be generalized to work with the empirical radial cumulative distribution function instead.}

An alternative is the slice tour \citep{sliceTour} which allows
distributions of points around the center of the data to be explored
using sections instead of projections. Instead of adjusting for effects
from projecting a high-dimensional volume, the solution with the slice
tour is to only highlight projected points if they are within a small
volume around the current projection plane, thus showing a local view of
the distribution. The slice tour is useful when there are large numbers
of observations or if there is concave structure in the data. Note that
only few methods are currently available for looking at sections of the
data in a systematic fashion, and further work is needed to fully
explore the potential of this type of visualization. In smaller datasets
we may also use interactive approaches such as linked brushing to
explore for local structures or conditional features.

@@

The authors gratefully acknowledge the support of the Australian
Research Council. The paper was written in \texttt{rmarkdown}
\citep{rmarkdown} using \texttt{knitr} \citep{knitr}. While the paper
title was originally derived from a pun, we thank the reviewer who
suggested that it could also be considered an acronym: BuRniNG SaGE:
Baseline Radial Non-Linear Geometric Scaling for Graphical Exploration.
